### 2021-03-27 newvariant.tex
updates

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\rho$.
For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
As our baseline scenario we assume $\rho = 1.5$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\rho=1.5$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
}
\label{fig:relspeed}
\end{figure}

@@

In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\RR_v(t)/\RR_w(t) = \rho$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ need not be constant, where
\begin{equation}
\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
\end{equation}

Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--D);
(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}E--H); and
(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}I--L).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_v(t)/\hat{\RR}_w(t)$ with the true ratio $\rho = \RR_v(t)/\RR_w(t)$.
For simplicity, we did not model non-pharmaceutical interventions;
instead, we let epidemic dynamics to be determined by susceptible depletion alone.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{Rtbias.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, E, I) Daily incidence of infection caused by the new variant and the wild type.
(B, F, J) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(C, G, K) True (purple) and estimated (orange) ratios between reproduction numbers of the new variant and the wild type over time.
(D, H, L) Relationship between estimated reproduction numbers on a log-log sacle.
The dashed lines represent the one-to-one line. 
The dotted lines represent the $\RR_v=\rho \RR_w$ line. 
Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (3.5 days).
Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6.5 days).
For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
}
\label{fig:Rtbias}
\end{figure}

When the assumed distribution matches the true distributions, the estimated ratio remains constant (\fref{Rtbias}C).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}G,K).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\RR_v(t)$ during the growth phase and under-estimate (or over-estimate) $\RR_v(t)$ during the decay phase (\fref{Rtbias}F,J), which further translates to biases in the estimated relative strength (\fref{Rtbias}G,K).

We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_v(t)$ against the estimated strength of the wild type $\hat{\RR}_w(t)$ on a log-log scale (\fref{Rtbias}D,H,L).
When generation-interval distributions are correctly specified, we obtain a straight line parallel to the one-to-one line. 
In this case, the slope corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}D).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.

>>>>

While relative speed is easier to measure, we often do not expect $\delta$ to be constant---this can depend on epidemic and demographic factors and can vary across time and space.
Instead, biological mechanisms that cause new variant to have higher strength better translate to assuming a constant value of relative strength $\rho$.
For example, if the new variant has a higher transmission rate, the ratio between the strength of the new variant and that of the wild type will remain constant ($=\rho$) under strength-like interventions that reduces the strength of both strains by a constant amount.
However, this is not necessarily true under speed-like interventions (e.g., contact tracing and self-isolation), which depends on time since infection: if the generation-interval distributions are different, the effectiveness of interventions will also differ.
For now, we consider consequences of assuming a constant relative strength $\rho$ and predicting the relative speed $\delta$---an assumption often made (either implicitly or explicitly) in studies that have tried to predict the future spread of the new variant, including when it will replace the wild type.
As our baseline scenario we assume $\rho = 1.61$, which is the value we obtain for $\delta=0.1\,\textrm{days}$, $\bar{G}_w = \bar{G}_v = 5\,\textrm{days}$, and $\kappa = 1/5$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{relspeed.pdf}
\caption{
\textbf{Relative speed of the new variant given assumed relative strength.}
(A) Estimated relative speed of the new variant $\hat{\delta}$ given strength of the wild type $\RR_w$ assuming identical generation-interval distributions.
(B) True relative strength of the new variant $\delta$ given strength of the wild type $\RR_w$ and ratio between the mean generation interval of the new variant $\bar{G}_v$ and that of the wild type $\bar{G}_w$..
(C) Difference between the true and the estimated relative speed of the new variant $\delta - \hat{\delta}$.
True relative speed of the new variant $\delta$ is calculated using $\rho=1.61$, $\bar{G}_w = 5\,\textrm{days}$, and $\kappa=1/5$.
}
\label{fig:relspeed}
\end{figure}

@@

In practice, epidemiological conditions change over time due to intervention measures and susceptible depletion effects.
Such changes can be characterized by ``time-varying'' or ``instantaneous'' reproduction numbers, $\RR(t)$ \citep{fraser2007estimating}.
Assuming that the generation-interval distribution does not change over time, the instantaneous reproduction numbers of the new variant and of the wild type can be estimated from their corresponding incidence curves---an approach popularized by \cite{cori2013new}:
\begin{equation}
\RR_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) g_x(\tau) \dtau}.
\end{equation}
Under strength-like intervention measures that reduce transmission rates by a constant amount, we expect ratios between true reproduction numbers to remain constant and correspond to the true relative strength: $\Ry{var}(t)/\Ry{wt}(t) = \rho$.
However, if the assumed generation-interval distribution $\hat{g}(\tau)$ is different from the true distribution, then the ratio between the estimated reproduction numbers $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ need not be constant, where
\begin{equation}
\hat{\RR}_x(t) = \frac{i_x(t)}{\int_0^\infty i_x(t-\tau) \hat{g}(\tau) \dtau}.
\end{equation}

Here, we simulate three different scenarios using a two strain renewal equation that assumes perfect cross-immunity (\fref{Rtbias}): 
(1) the wild type and the variant have the same generation-interval distributions, which match the assumed distribution (\fref{Rtbias}A--C);
(2) the assumed distribution matches the generation-interval distribution of the wild type but has a longer mean than the mean generation interval of the variant (\fref{Rtbias}D--F); and
(3) the assumed distribution matches the generation-interval distribution of the wild type but has a shorter mean than the mean generation interval of the variant (\fref{Rtbias}G--I).
Then, we compare the estimated ratio $\hat{\rho}(t) = \hat{\RR}_{\textrm{var}}(t)/\hat{\RR}_{\textrm{wt}}(t)$ with the true ratio $\rho = \Ry{var}(t)/\Ry{wt}(t)$.
In order to reflect smooth changes in $\RR$ driven by non-pharmaceutical interventions, we model $\Ry{wt}(t)$ phenomenologically and let $\Ry{var}(t) = \rho \Ry{wt}(t)$.

\begin{figure}[!th]
\includegraphics[width=\textwidth]{Rtbias.pdf}
\caption{
\textbf{Estimates of relative strength over time under different scenarios.}
(A, D, G) True (solid lines) and estimated (dashed lines) reproduction numbers of the new variant and the wild type over time.
(B, E, H) True (purple, solid) and estimated (orange, dashed) ratios between reproduction numbers of the new variant and the wild type over time.
(C, F, I) True (purple, solid) and estimated (orange, dashed) relationships between estimated reproduction numbers on a log-log sacle.
Top row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type and the variant.
Middle row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (4 days).
Bottom row: the assumed mean generation-interval (5 days) is equal to the mean generation-interval of the wild type but is longer than that of the varant (6 days).
For all simulations, squared coefficient of variation in generation intervals is assumed to equal $\kappa = 1/5$.
We assume $\Ry{wt}(t)=2.5$ before $t=15\,\textrm{days}$ to reflect unmitigated spread and let $\Ry{wt}(t)=2.5 (1+0.7\cos((t-15)/10))/1.7$ after to model smooth introduction and lifting of non-pharmaceutical interventions.
}
\label{fig:Rtbias}
\end{figure}

When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Ry{var}(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).

We can assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ on a log-log scale (\fref{Rtbias}C,F,I).
When generation-interval distributions are correctly specified, we obtain a straight line with a slope of 1. 
In this case, the intercept corresponds to the logarithm of the relative strength $\log(\rho)$ (\fref{Rtbias}C).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L) on a log-log plot.
Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
